---
title: 'Formative Practical Report: Summarising Multivariate Data and PCA'
author: "Jack Young"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nclSLR)
library(ggplot2)
library(tidyr)
```

## Numerical and Graphical summaries

We start by selecting the `airpollution` dataset from our package. We can draw immediate insights using the summary function - as shown below, this shows us all of the variables in our data as well as some summary statistics for each variable:

```{r intro,message=FALSE}
summary(airpollution)
```

We can also visualize this information using histogram plots for each variable - 
these give us a little more insight as to how the values are distributed:
```{r boxplots, message=FALSE}
ggplot(gather(airpollution), aes(value)) +
  geom_histogram(bins = 10) +
  facet_wrap(~key, scales = 'free_x')

```

As above, we see a variety of distributions present in our data. For example,
some seem to roughly follow a normal distribution, whereas others do not, such
as the 'PM2' variable which appears to follow an exponential distribution.

We can check the dimensions of the data like so:
```{r dims}
dim(airpollution)
```

In the context of this dataset, this means that we have 80 observations, and 11
variables in total. We could produce a scatterplot matrix for our data, but this
may be impractical in this case given the large number of variables present. A
good alternative for this is to use a correlation heatmap instead, which
colour-codes variable pairs based on their correlation coefficient:

```{r pairs}
cor(airpollution)
corrplot::corrplot(cor(airpollution))
```

For example, we can see here that there is a fairly strong correlation between
variables 'PERWH' and 'NONPOOR'.
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
